---
title: "MesoNet reproduction — statistical analysis"
subtitle: "Video-clustered GLMM, DeLong AUC comparisons, and bootstrap CIs (Task 6)"
format:
  html:
    toc: true
    embed-resources: true
    theme: cosmo
execute:
  echo: false
  warning: false
---

```{r setup}
library(readr); library(dplyr); library(knitr)
fixed  <- read_csv("output/glmm_fixed_effects.csv", show_col_types = FALSE)
ranef  <- read_csv("output/glmm_ranef_sd.csv", show_col_types = FALSE)
cells  <- read_csv("output/emmeans_cells.csv", show_col_types = FALSE)
contr  <- read_csv("output/emmeans_contrasts.csv", show_col_types = FALSE)
delong <- read_csv("output/delong.csv", show_col_types = FALSE)
cis    <- read_csv("output/auc_cis.csv", show_col_types = FALSE)
fmt <- function(x, d = 3) formatC(x, format = "f", digits = d)
```

## What this analysis is for

The repo's README reports accuracies and AUCs as mean ± std over three training seeds. Those
summaries treat every frame as independent — but 20 frames from the same source video are
strongly correlated, so naive per-frame uncertainty is overconfident. The headline analysis here
is therefore a **mixed-effects logistic regression (GLMM) with a random intercept per source
video** (and per seed), fit on the 67,200 in-domain FaceForensics++ test predictions
(2 models × 2 methods × 3 seeds × 5,600 frames):

`correct ~ model * method + (1 | video_id) + (1 | seed)` — references:
`model = meso_inception4`, `method = Face2Face`.

**Data provenance:** one row per evaluated image, exported by `scripts/export_for_r.py` from the
saved per-run probability files; regenerate with that script, then `Rscript models.R`, then
`quarto render report.qmd`.

## GLMM — fixed effects (odds ratios)

```{r fixed}
kable(fixed |> mutate(across(where(is.numeric), ~fmt(.x))), align = "l",
      col.names = c("Term", "Odds ratio", "CI low", "CI high", "p"))
```

```{r fixed-values}
or_row <- function(t) fixed |> filter(term == t)
m4  <- or_row("modelmeso4")
dfk <- or_row("methodDeepfakes")
```

Reading the table (odds ratios for a frame being classified **correctly**):

- **Meso-4 vs MesoInception-4, Face2Face** (the reference cell): OR `r fmt(m4$estimate, 2)`
  (95% CI `r fmt(m4$conf.low, 2)`–`r fmt(m4$conf.high, 2)`) — spans 1, the two architectures are
  statistically indistinguishable there.
- **Meso-4 vs MesoInception-4, Deepfakes**: the interaction is significant. With Meso-4 in the
  numerator, the emmeans contrast gives **OR ≈ 1.23 (95% CI 1.13–1.35), p < 0.001** — Meso-4's
  threshold-0.5 advantage on Deepfakes is real, not seed noise. Crucially, the DeLong section
  shows the models' *AUCs* are essentially equal with seed-flipping sign, and the README's
  threshold-tuning result shows the accuracies converge once each model gets its own validated
  threshold. Together: a **statistically significant calibration advantage** for Meso-4 at the
  default threshold — not better ranking.
- **Deepfakes vs Face2Face** (for MesoInception-4): OR `r fmt(dfk$estimate, 2)`
  (CI `r fmt(dfk$conf.low, 2)`–`r fmt(dfk$conf.high, 2)`) — at the 0.5-threshold decision level
  Deepfakes is *harder* for MesoInception-4, reversing the paper's difficulty ordering; for
  Meso-4 the interaction cancels it (net OR ≈ 1.03). The paper's DF-easier-than-F2F ordering
  shows up in AUC (0.983 vs 0.972), not in thresholded accuracy — another calibration effect.
- **Video-level variance dominates:** the `video_id` random-effect SD is
  `r fmt(ranef$sd[ranef$group == "video_id"], 2)` on the log-odds scale vs
  `r fmt(ranef$sd[ranef$group == "seed"], 2)` for `seed` — *which video a frame comes from
  matters two orders of magnitude more than the training seed.* Some videos are simply hard:
  this is exactly why the clustered analysis is the rigorous one.

## Per-cell accuracies and the model contrast (emmeans)

```{r cells}
kable(cells |> mutate(across(where(is.numeric), ~fmt(.x))), align = "l")
```

```{r contrasts}
kable(contr |> mutate(across(where(is.numeric), ~fmt(.x))), align = "l")
```

Note the per-cell probabilities (0.97–0.98) are **conditional on an average video** (random
effect at zero) — higher than the raw marginal frame accuracies (0.91–0.93) because a minority
of hard videos drags the marginal mean down; the contrast between cells is the meaningful part.
The model contrast is n.s. on Face2Face (Meso-4-numerator OR ≈ 0.94, CI 0.86–1.03) but
significant on Deepfakes (Meso-4-numerator OR ≈ 1.23, CI 1.13–1.35, p < 0.001; the table above
stores the reciprocal MesoInception-4/Meso-4 direction) — the calibration reading above, not a
ranking difference.

## DeLong tests — are the two models' AUCs different?

Paired DeLong comparisons on identical test frames, per method and seed:

```{r delong}
kable(delong |> mutate(across(where(is.numeric), ~fmt(.x, 4))), align = "l",
      col.names = c("Method", "Seed", "AUC Meso-4", "AUC MesoInception-4", "Δ", "p (DeLong)"))
```

*Caveat:* DeLong treats frames as independent, which video clustering violates — these p-values
are anti-conservative and reported as a secondary, protocol-standard check rather than the
headline inference. Several Deepfakes comparisons are nominally significant (5,600 frames give
power to detect |Δ| ≈ 0.003), **but the sign flips across seeds** (+, +, −), and every
Face2Face comparison is n.s. — there is no consistent AUC superiority for either model, which is
precisely why the significant GLMM Deepfakes contrast should be read as calibration, not
ranking.

## Bootstrap CIs for every reported AUC

2,000 stratified bootstrap replicates per AUC. `excludes_chance` marks CIs that do not contain
0.5:

```{r cis}
kable(cis |> arrange(desc(in_domain), train_source, eval_dataset, seed) |>
        mutate(across(where(is.numeric), ~fmt(.x))), align = "l")
```

```{r cis-values}
cross <- cis |> filter(!in_domain, eval_dataset %in% c("openforensics", "faces140k"))
n_below <- sum(cross$ci_hi < 0.5)
```

The cross-dataset story, now with uncertainty: of the `r nrow(cross)` frozen-checkpoint
evaluations on truly unseen datasets (OpenForensics, 140k), **`r n_below` have an entire 95% CI
below 0.5** — the mild ranking inversion in the README is statistically supported per seed, not
an artifact of point estimates. The cross-method control (Face2Face) CIs sit well above 0.5,
confirming genuine (if degraded) transfer within the preprocessing domain.

## Limitations

- **Three seeds** is enough to bound run-to-run variance informally but not to estimate the seed
  random effect precisely; its SD estimate above is unstable by construction.
- DeLong p-values ignore frame clustering (noted above); the GLMM is the rigorous result.
- The GLMM models the 0.5-threshold decision; it says nothing about calibration (covered by the
  threshold-tuning analysis in the README).
- All FaceForensics++ results are c23 compression, per-image unless stated; cross-dataset rows
  reuse the frozen Deepfakes-trained Meso-4.
```
